Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent

### 2023-06-27 manuscript/main.tex
bretzel

We present a new pangenome graph layout algorithm which uses path-guided stochastic gradient descent %\citep{Zheng2019}
(PG-SGD) to move pairs of nodes in a HOGWILD! parallel manner. The algorithm aims to adjust the graph layout to best reflect the path nucleotide positions.
We exemplify how our implementation is able to produce human readable layouts. Furthermore, the 1D PG-SGD already is a key step in pangenome graph construciton \citep{Liao2023, Garrison2023} and analysis \citep{Guarracino2022, Guarracino2023}.

@@

\paragraph{The HOGHWILD! path-guided stochastic gradient descent algorithm}
\FIXME{SPECIFY THAT THE IMPLEMENTATION IS IN ODGI AND CITE IT}
Our algorithm is inspired by Zheng et al. ((\FIXME{CITATION} \url{https://doi.org/hkv4}) and it is applicable in any number of dimensions.
In this work, we describe the results in 1 dimension (1D) and 2 dimensions (2D).
In our implementation, the algorithm moves a single pair of nodes at a time, minimizing the disparity between the layout distance of a node pair and the actual nucleotide distance of a path traversing these nodes.
Our path index enables efficient navigation of path positions by updating nodes on the fly, instead of keeping all pairwise distances of all nodes in memory.
This allows the PG-SGD to be applied to gigabase-scale pangenome graphs. \\
Our multithreaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
\begin{itemize}
    \item The first node Xi of a pair is a uniform path step pick from all nodes.
    \item The second node Xj of a pair is sampled from the same path following a Zipfian distribution.
    \item The path nucleotide distance of the nodes in the pair guides the actual layout distance dij update of these nodes.
    \item The magnitude r of the update depends on the current learning rate of the SGD.
\end{itemize}
%The exploration of the path-guided SGD parameter space is in progress, in order to get the best layout as quickly as possible.
We describe in algorithm~\ref{alg:pg_sgd} how the PG-SGD works on a single thread.
We can have several worker threads but only one checker thread!

@@

    We presented the Path-Guided Stochastic Gradient Descent (PG-SGD), a layout algorithm for pangenome graphs that
    leverages the inherent biological information available within the graph's represented genomic sequences.
    Our implementation efficiently computes the layout of pangenome graphs representing thousands of individuals at whole-genome scale.
    \FIXME{Difference to other existing methods, are there possible improvements of our method possible?}
    \FIXME{Discuss performance eval. @Jiajie + Niklas: What about going GPU?}
    We implemented the PG-SGD algorithm to generate layouts in 1 and 2 dimensions, but the algorithm can be extended in any number of dimensions.
    Graph visualization is key for understanding genome variations and the large-scale layouts produced by the PG-SGD algorithm offer an unprecedented 
    high-level perspective on variation in pangenomes. These layouts are critical for pangenome graph building (\FIXME{CITE PGG and HPRC MAIN PAPER}) and 
    making new biological discoveries (\FIXME{CITE ACROCENTRIC PAPER}).
    %The graph simplification pipeline smoothxg runs POA for each block of paths that are co-linear within each seqwish induced variation graph.
    %A prerequisite is that the graph nodes are sorted according to their occurrence in the graph's embedded paths.
    %Our 1D path-guided SGD algorithm is designed to provide this kind of sort.
    %Already, the 1D PG-SGD is a key step in the PanGenome Graph Building (PGGB) pipeline that we successfully applied to build the first draft human pangenome reference (Liao et al., bioRxiv 2022).
    However, there remains a gap in interactive and scalable solutions that merge layouts of large pangenome graph with annotation.
    Our PG-SGD algorithm will be the foundation of new pangenome graph browsers for exploring meaningful graph layouts.
    Futhermore, projecting a graph in a low number of dimensions, but still preserving its biologically relevant information, enable the application of 
    algorithms that use the graph layout to look at variant classification.
    Our future research involve exploiting these learned projections to detect structural variations and assembly errors.
    \FIXME{PROTEIN?}

>>>>

	This paper describes a new pangenome graph layout algorithm which uses path-guided stochastic gradient descent %\citep{Zheng2019}
	(PG-SGD) to move pairs of nodes in a HOGWILD! parallel manner. The algorithm aims to adjust the graph layout to best reflect the path nucleotide positions. In theory, our algorithm is applicable in any number of dimensions. Here,
	we exemplify how our implementation is able to produce 1D and 2D human readable layouts. Already, the 1D PG-SGD is a key step in pangenome graph construction \citep{Liao2023, Garrison2023} and analysis \citep{Guarracino2022, Guarracino2023}.

@@

	In this work, we describe the results in 1 dimension (1D) and 2 dimensions (2D). In this section we will describe. Then we will. And then. \\

	\FIXME{Do we need a formal definition of a variation graph like in the seqwish paper?}

	Our HOGHWILD! path-guided stochastic gradient descent algorithm is inspired by \cite{Zheng2019} and is theoretically applicable in any number of dimensions. 
	We implemented the algorithm in ODGI \citep{Guarracino2022}, making use of its already proven interface to pangenome graphs in the standard graphical fragment assebmly (GFA) format (\url{https://github.com/GFA-spec/GFA-spec}).
	In our implementation, the algorithm moves a single pair of nodes at a time, minimizing the disparity between the layout distance of a node pair and the actual nucleotide distance of a path traversing these nodes.
	Our path index enables efficient navigation of path positions by updating nodes on the fly, instead of keeping all pairwise distances of all nodes in memory.
	This allows the PG-SGD to be applied to gigabase-scale pangenome graphs. \\
	Our multithreaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
	\begin{itemize}
	    \item The first node Xi of a pair is a uniform path step pick from all nodes.
	    \item The second node Xj of a pair is sampled from the same path following a Zipfian distribution.
	    \item The path nucleotide distance of the nodes in the pair guides the actual layout distance dij update of these nodes.
	    \item The magnitude r of the update depends on the current learning rate of the SGD.
	\end{itemize}
	%The exploration of the path-guided SGD parameter space is in progress, in order to get the best layout as quickly as possible.
	We describe in algorithm~\ref{alg:pg_sgd} how the PG-SGD works on a single thread.
	We can have several worker threads but only one checker thread!

	\subsection{general algorithm}
	\subsection{path index}
	\subsection{parallelization in detail}
	\subsection{randomization}
	ZIPF!
	\subsection{annealing step-size $\eta$}
	\subsection{convergence / stop criteria, iterations, term updates}
	\subsection{cooling || flip}
	\subsection{from layout to 1D node order}

@@

	We have presented the PG-SGD layout algorithm for pangenome graphs that leverages the inherent biological information available within the graph's represented genomic sequences.
	Our implementation efficiently computes the layout of pangenome graphs representing thousands of individuals at whole-genome scale.
	\FIXME{Difference to other existing methods, are there possible improvements of our method possible?}
	\FIXME{Discuss performance eval. @Jiajie + Niklas: What about going GPU?}
	We implemented the PG-SGD algorithm to generate layouts in 1D and 2D, but the algorithm can in princible be extended to any number of dimensions.
	Graph visualization is key for understanding genome variations and the large-scale layouts produced by the PG-SGD algorithm offer an unprecedented 
	high-level perspective on variation in pangenomes. These layouts are critical for pangenome graph building (\FIXME{CITE PGG and HPRC MAIN PAPER}) and 
	making new biological discoveries (\FIXME{CITE ACROCENTRIC PAPER}).

	Problem space: We take a look at all paths (FULL HPRC GRAPH ~8 BILLION STEPS ACROSS ~34 THOUSAND PATHS) and not at all nodes (FULL HPRC GRAPH ~100 MILLION). So we have a very large problem space compared to just looking at the nodes.
	%The graph simplification pipeline smoothxg runs POA for each block of paths that are co-linear within each seqwish induced variation graph.
	%A prerequisite is that the graph nodes are sorted according to their occurrence in the graph's embedded paths.
	%Our 1D path-guided SGD algorithm is designed to provide this kind of sort.
	%Already, the 1D PG-SGD is a key step in the PanGenome Graph Building (PGGB) pipeline that we successfully applied to build the first draft human pangenome reference (Liao et al., bioRxiv 2022).

	However, there remains a gap in interactive and scalable solutions that merge layouts of large pangenome graph with annotation.
	Our PG-SGD algorithm will be the foundation of new pangenome graph browsers for exploring meaningful graph layouts.
	\FIXME{2D: waragraph + gfaestus for interactive visualization?}
	Futhermore, projecting a graph in a low number of dimensions, but still preserving its biologically relevant information, enable the application of algorithms that use the graph layout to look at variant classification.
	Our future research involve exploiting these learned projections to detect structural variations and assembly errors.
	\FIXME{PROTEIN?}

### 2023-06-27 manuscript/pg-sgd_pseudocode.tex
bretzel

\begin{algorithm}
	\pgsgd{\textbf{(}$\mathcal{V}$\textbf{)}}{
		\textbf{input:} variation graph $\mathcal{V} = (\mathcal{N}, \mathcal{E}, \mathcal{P})$ \\
		%https://tex.stackexchange.com/questions/22643/how-to-write-letters-in-bold-in-the-math-mode
		\textbf{output:} $k$-dimensional layout $\mathcal{L}$ with $n$ nodes \\
		$\mathcal{XP}$ $\gets \PathIndex(\mathcal{V})$ \tcp{index for efficient path position lookup}
		%\boldsymbol{$Z$} $\gets ZipfZetas(G,P)$ \\ %@Andrea is this correct like this?
		$\mathcal{L}$ $\gets \RandomLayout(n, k)$ \tcp{only 2D PG-SGD has a random layout initialization, 1D starts with the current node order}
		$\mathcal{Z} \gets \InitZipf(\mathcal{V},\mathcal{XP})$ \tcp{generate Zipfian distribution}
		% atomic positions initialization?
		% TODO for somplicity reasons I would only describe the 2D one here
		\for{$\eta$ $in$ $annealing$ $schedule$}{ %our "schedule" actually is the number paths.... we should specify this I would say
			\each{$planned$ $term$ $updates$} { % I think we can remove i<j, because we don't care about that
				$ps_a \gets \Unif(\mathcal{XP})$ \tcp{uniform path step from all steps}
				$path \gets \PathId(ps_a,\mathcal{XP})$ \tcp{the path id of the sampled step}
				\uIf{$(coolling$ $||$ $flip)$} {
					$ps_b \gets \Unif(\StepCount(path, \mathcal{XP}))$ \tcp{uniform path step from specific path}
					$p_b \gets \StepPos(ps_b)$ \tcp{nucleotide position of step}
				} \Else {
					$ps_b \gets \Zipf(path)$ \tcp{Zipfian path step from specific path}
				}
				$p_a \gets \StepPos(ps_a)$ \tcp{nucleotide position of step}
				$p_b \gets \StepPos(ps_b)$ \tcp{nucleotide position of step}
				$nd \gets ||p_a - p_b||$ \tcp{calculate nucleotide distance} %mag is nx
				$ld \gets ||\mathcal{L}_{a}-\mathcal{L}_b||$ \tcp{calculate layout distance}
				$w_{ab} \gets \frac{1.0}{nd}$ \tcp{term weight} 
				$\mu \gets w_{ab}\eta$ \tcp{learning rate}  % current learning rate is given by term weight and step size
				\IF{$\mu>1$} {
				$\mu \gets 1$
				}
				$\delta \gets \mu \cdot \frac{ld -nd}{2}$ \tcp{the actual delta}
				\uIf{$\delta <= 0$} {
					$STOP$ \tcp{we can't optimize it any better, so we stop here}
				}
				% TODO stop early?
				% TODO potentially store new delta max?
				$r \gets \delta - ld$ \tcp{size of the term update}
				$\mathcal{L}_a \gets \mathcal{L}_a - r\cdot ld$ \tcp{update the 1. term}
				$\mathcal{L}_b \gets \mathcal{L}_b - r\cdot ld$ \tcp{update the 2. term}
			}
		}
	}
	\caption{Pseudocode of the PG-SGD algorithm. Explain $\eta$. Explain $PathIndex$ in the text. Explain $ZipfZetas$ in the text. Explain $RandomLayout$ in the text. Explain annealing schedule in the text or rephrase this wording.}
	\label{alg:pg_sgd}
\end{algorithm}

>>>>

\begin{algorithm}
	\pgsgd{\textbf{(}$\mathcal{V}$\textbf{)}}{
		\textbf{input:} variation graph $\mathcal{V} = (\mathcal{N}, \mathcal{E}, \mathcal{P})$ \\
		%https://tex.stackexchange.com/questions/22643/how-to-write-letters-in-bold-in-the-math-mode
		\textbf{output:} $k$-dimensional layout $\mathcal{L}$ with $n$ nodes \\
		$\mathcal{XP}$ $\gets \PathIndex(\mathcal{V})$ \tcp{index for efficient path position lookup}
		%\boldsymbol{$Z$} $\gets ZipfZetas(G,P)$ \\ %@Andrea is this correct like this?
		$\mathcal{L}$ $\gets \RandomLayout(n, k)$ \tcp{only 2D PG-SGD has a random layout initialization, 1D starts with the current node order}
		$\mathcal{Z} \gets \InitZipf(\mathcal{V},\mathcal{XP})$ \tcp{generate Zipfian distribution}
		% atomic positions initialization?
		% TODO for somplicity reasons I would only describe the 2D one here
		\for{$\eta$ $in$ $annealing$ $schedule$}{ %our "schedule" actually is the number paths.... we should specify this I would say
			\each{$planned$ $term$ $update$} { % I think we can remove i<j, because we don't care about that
				$ps_a \gets \Unif(\mathcal{XP})$ \tcp{uniform path step from all steps}
				$path \gets \PathId(ps_a,\mathcal{XP})$ \tcp{the path id of the sampled step}
				\uIf{$(cooling$ $||$ $flip)$} {
					$ps_b \gets \Unif(\StepCount(path, \mathcal{XP}))$ \tcp{uniform path step from specific path}
					$p_b \gets \StepPos(ps_b)$ \tcp{nucleotide position of step}
				} \Else {
					$ps_b \gets \Zipf(path)$ \tcp{Zipfian path step from specific path}
				}
				$p_a \gets \StepPos(ps_a)$ \tcp{nucleotide position of step}
				$p_b \gets \StepPos(ps_b)$ \tcp{nucleotide position of step}
				$nd \gets ||p_a - p_b||$ \tcp{calculate nucleotide distance} %mag is nx
				$ld \gets ||\mathcal{L}_{a}-\mathcal{L}_b||$ \tcp{calculate layout distance}
				$w_{ab} \gets \frac{1.0}{nd}$ \tcp{term weight} 
				$\mu \gets w_{ab}\eta$ \tcp{learning rate}  % current learning rate is given by term weight and step size
				\IF{$\mu>1$} {
				$\mu \gets 1$
				}
				$\delta \gets \mu \cdot \frac{ld -nd}{2}$ \tcp{the actual delta}
				\uIf{$\delta <= 0$} {
					$STOP$ \tcp{we can't optimize it any better, so we stop here}
				}
				% TODO stop early?
				% TODO potentially store new delta max?
				$r \gets \delta - ld$ \tcp{size of the term update}
				$\mathcal{L}_a \gets \mathcal{L}_a - r\cdot ld$ \tcp{update the 1. term}
				$\mathcal{L}_b \gets \mathcal{L}_b - r\cdot ld$ \tcp{update the 2. term}
			}
		}
	}
	\caption{Pseudocode of the PG-SGD algorithm. Explain $\eta$. Explain $PathIndex$ in the text. Explain $InitZipf$ in the text. Explain $RandomLayout$ in the text. Explain annealing schedule in the text or rephrase this wording.}
	\label{alg:pg_sgd}
\end{algorithm}
